---
title: "Comparative Analysis: NY State Vs. NYC"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

&nbsp; 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(data.table)
library(janitor)
library(ggridges)
library(patchwork)
library(readxl)
library("writexl")
library(rvest)
library(httr)
library(dplyr)
library(plotly)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


```{r data_import, message=FALSE}
data = read_csv(file = "./data/master.csv") %>%
  janitor::clean_names()
```

From the master spreadsheet, we selected the columns and rows we will be using to analyze induced abortions by covariate categories. We first used `pivot_longer` to reformat the data and used `plotly` to create interactive plots. 

&nbsp; 

### NY vs NYC Age
***

```{r}
age_total = 
  data %>%
  select(1:7) %>%
  slice(1:2)

age_total %>%
  pivot_longer(
    age_less_20:age_plus_40,
    names_to = "age", 
    names_prefix = "age_",
    values_to = "abortion"
  ) %>% 
  mutate(
    age = replace(age, age == "less_20", "<20"),
    age = replace(age, age == "20_24", "20-24"),
    age = replace(age, age == "25_29", "25-29"),
    age = replace(age, age == "30_34", "30-34"),
    age = replace(age, age == "35_39", "35-39"),
    age = replace(age, age == "plus_40", "40+"),
    age = factor(age)
  ) %>% 
  plot_ly(x = ~age, y = ~abortion, color = ~borough, type = "bar", colors = "viridis") %>%   
 layout(
    title = 'Induced Abortions in NY State & NYC by Age', 
    xaxis = list(title = "Age (Years)"),
    yaxis = list(title = 'Abortions per 1,000 Live Births'))

age_total
```

From the bar plots, it is evident that those in New York City have higher abortion rates compared to those in New York State across all age categories. Rates seem to decrease as age categories increase for New York City; however, for New York State, rates decrease from the <20 to 35-39 age categories, and increase among those who are 40+. Rates were highest among those who were <20 years for both New York City and New York State. Those <20 years in New York State had the highest rates across all categories of 1491.4 abortions per 1000 live births.

&nbsp; 

### NY vs. NYC Race
***

```{r}
race_total = 
  data %>%
  select(1,9:12) %>%
  slice(1:2)

race_total %>%
  pivot_longer(
    nh_white_only_ratio:h_total,
    names_to = "race", 
    values_to = "abortion"
  ) %>%
  mutate(
    race = replace(race, race == "nh_white_only_ratio", "Non-Hispanic White Only"),
    race = replace(race, race == "nh_black_only_ratio", "Non-Hispanic Black Only"),
    race = replace(race, race == "nh_other_ratio", "Non-Hispanic Other"),
    race = replace(race, race == "h_total", "Hispanic"),
    race = factor(race)
    )%>% 
  plot_ly(x = ~race, y = ~abortion, color = ~borough, type = "bar", colors = "viridis") %>%   
   layout(
    title = "Induced Abortions in NY State & NYC by Race/Ethnicity", 
    xaxis = list(title = "Race"),
    yaxis = list(title = "Abortions per 1,000 Live Births"))

race_total
```

It is evident that those who identified as Non-Hispanic Black Only had the highest abortion rates across all boroughs in NYC. Among those who identified as Non-Hispanic Black, rates were highest in Manhattan with 851.7 induced abortions per 1,000 live births. Those who identified as Non-Hispanic White and Non-Hispanic Other had some of the lowest abortion rates in NYC. The lowest rates across all boroughs and race groups was for those who identified as Non-Hispanic White only and lived in Brooklyn with a rates of 142.9 induced abortions per 1,000 live births. 

&nbsp; 

### NY vs. NYC Financial Plans
***

```{r}
finance_plan_total= 
  data %>%
  select(1,15:18) %>%
  slice(1:2)

finance_plan_total %>%  
  pivot_longer(
    medicaid:not_stated,
    names_to = "finance_plan", 
    values_to = "abortion"
  ) %>%
  mutate(
    finance_plan = replace(finance_plan, finance_plan == "medicaid", "Medicaid"),
    finance_plan = replace(finance_plan, finance_plan == "self_pay", "Self Pay"),
    finance_plan = replace(finance_plan, finance_plan == "other_insurance", "Other Insurance"),
    finance_plan = replace(finance_plan, finance_plan == "not_stated", "Not Stated"),
    finance_plan = factor(finance_plan)
    ) %>% 
  plot_ly(x = ~finance_plan, y = ~abortion, color = ~borough, type = "bar", colors = "viridis") %>%   
  layout(
    title = "Induced Abortions in NY State by Financial Plan Status", 
    xaxis = list(title = "Type of Financial Plan"),
    yaxis = list(title = "Abortions per 1,000 Live Births"))
```

Rates were highest for those with Other Insurance in both New York City and New York State and was higher in New York City with 19.92k induced abortions per 1000 live births. Rates were lowest for those with Medicaid in both New York City and New York State and was lower for those in New York State with 325.66 induced abortions per 1000 live births. 

