---
title: "Comparative Statistical Analysis"
output: 
  html_document: 
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

&nbsp; 

```{r setup, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r data, message=FALSE, include=FALSE}
abortion_data = read_csv("data/master.csv")
```

```{r data_NY, message=FALSE, include=FALSE}
NY_year = read_csv("data/NY_year.csv")
```

## Abortions Rates Comparing NYC and NY State

&nbsp; 

### Proportion T-test
***

**Do overall abortion rates differ between NYC and NY state?**

**$H_0$**: The abortion rate per 1000 births is the same between NYC and NY state.

**$H_1$**: The abortion rate per 1000 births differs between NYC and NY state.

```{r tidy_totals}
# Tidy data
abortion_totals =
  abortion_data %>%
  select(borough, total) %>%
  head(2) %>%
  mutate(
    n = total,
    total = 1000)

abortion_totals %>% 
  select (-total) %>%
  knitr::kable(caption = "Proportion T-Test for NYC and NY State", col.names = c("Region", "Abortion Rate per 1000 births"), align = c('l', 'c'), format = "simple")
```

```{r prop}
# Run prop t-test
prop.test(abortion_totals$n, abortion_totals$total,
          p = NULL, alternative = "two.sided")
```

**Inference:**

- From the above results, the p-value is **less than the 5% level of significance.** Therefore, we **reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that the abortion rate per 1000 births differs between NYC and NY state.