---
title: "New York State Exploratory Analysis"
---

```{r setup, message = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(ggridges)
library(patchwork)
library(plotly)

knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r merged data}
merged = read_csv("data/master.csv") %>% 
  janitor::clean_names()
```

### Total

```{r NY year}
NY_year = read_csv("data/NY_year.csv")

plot_borough_year = 
  NY_year %>% 
  drop_na() %>% 
  mutate(borough = fct_reorder(borough, total)) %>% 
  plot_ly(
    y = ~total, x = ~year, color = ~borough, 
    type = "scatter", mode = "line", colors = "viridis") %>% 
   layout(
     title = 'Abortion Rates from 2014-2019 by Regions in NYS', 
     xaxis = list(title = "Year"),
     yaxis = list(title = 'Abortions per 1,000 Live Births'))

plot_borough_year

table_nys =
  NY_year %>% 
  mutate(borough = fct_reorder(borough, total)) %>%
  select(year, borough, total) %>% 
  pivot_wider(
    names_from = "borough",
    values_from = "total") %>% 
  knitr::kable(caption = "Abortion Rates in New York State by Region (2014-2019)", col.names = c("Year", "Central New York", "Eastern Adirondacks", "Finger Lakes", "Hudson Valley", "Long Island", "New York City", "Saratoga", "Westchester/ Rockland", "Western Adirondacks", "Western New York"), align = 'lcccccccccc', format = "simple")

table_nys
  
```

From the plot above, it is evident that, on average, New York City has higher rates of abortions compared to the rest of the regions within New York State. We also wanted to take a look at what's happening at the individual county-level. To find out more about county-level trends, head over to the [dashboard]() that we created! 

#### Age 

```{r age nys}
age_plot_nys =
  merged %>% 
  select(region, starts_with("age")) %>% 
  pivot_longer(
    age_less_20:age_plus_40,
    names_to = "age", 
    values_to = "abortion",
    names_prefix = "age_"
  ) %>% 
  mutate(
    age = replace(age, age == "less_20", "<20"),
    age = replace(age, age == "20_24", "20-24"),
    age = replace(age, age == "25_29", "25-29"),
    age = replace(age, age == "30_34", "30-34"),
    age = replace(age, age == "35_39", "35-39"),
    age = replace(age, age == "plus_40", "40+"),
    age = factor(age)
  ) %>%
  group_by(region,age)%>%
  summarise(total = sum(abortion)) %>% 
  plot_ly(
    x = ~age, y = ~total, color = ~region, 
    type = "bar", colors = "viridis") %>% 
  layout(
    title = 'Induced Abortions in NY State by Age', 
    xaxis = list(title = "Age (Years)"),
    yaxis = list(title = 'Abortions per 1,000 Live Births'))

age_plot_nys
```

From the above plot, the highest abortion rate per 1000 live births is generally among those that are below 20 years of age when comparing different categories of age. We see a decline in induced abortion rates among older people across most regions within New York State, but some rates start to increase for age categories 35-39 and 40+. The highest abortion rate across all categories is among those in New York City and less than 20 years old with 7,677.5 induced abortions per 1000 live births. The second highest abortion rates across all categories is in the Capital/Saratoga region for those less than 20 years old with 7344.9 induced abortions per 1000 live births. While New York city has higher abortion rates across most age categories, we see that among those who are 40+, higher rates of abortion are found among those in the Finger Lakes, Capital/Saratoga, and Central New York regions when compared to New York City.

#### Race/Ethnicity

```{r}
race_plot_nys =
  merged %>% 
  select(region, nh_white_only_ratio:h_total) %>%
  pivot_longer(
    nh_white_only_ratio:h_total,
    names_to = "race", 
    values_to = "abortion"
    ) %>%
  mutate(
    race = replace(race, race == "nh_white_only_ratio", "Non-Hispanic White Only"),
    race = replace(race, race == "nh_black_only_ratio", "Non-Hispanic Black Only"),
    race = replace(race, race == "nh_other_ratio", "Non-Hispanic Other"),
    race = replace(race, race == "h_total", "Hispanic"),
    race = factor(race)
    ) %>%
  group_by(region,race)%>%
  summarise(total = sum(abortion)) %>% 
  plot_ly(
    x = ~race, y = ~total, color = ~region, 
    type = "bar", colors = "viridis") %>% 
  layout(
    title = "Induced Abortions in NY State by Race/Ethnicity", 
    xaxis = list(title = "Race"),
    yaxis = list(title = "Abortions per 1,000 Live Births"))

race_plot_nys
```

The highest rate of abortion is among those residing in Hudson Valley and identify as Non-Hispanic Black with 4638.1 induced abortions per 1000 live births. Non-Hispanic Blacks residing in New York City have the second highest rate as well with 4387.6 induced abortions per 1000 live births. There were many regions that are missing data in the plot because the data is missing for some counties in the original data set. For the statistical analysis, we will conduct statistical tests by county to get finer comparisons with more datapoints. 

#### Finacial Plan Status

```{r}
finance_plot_nys= 
  merged %>%
  select(region, medicaid:not_stated) %>%
  pivot_longer(
    medicaid:not_stated,
    names_to = "finance_plan", 
    values_to = "abortion"
  ) %>% 
  mutate(
    finance_plan = replace(finance_plan, finance_plan == "medicaid", "Medicaid"),
    finance_plan = replace(finance_plan, finance_plan == "self_pay", "Self Pay"),
    finance_plan = replace(finance_plan, finance_plan == "other_insurance", "Other Insurance"),
    finance_plan = replace(finance_plan, finance_plan == "not_stated", "Not Stated"),
    finance_plan = factor(finance_plan)
    ) %>%
  group_by(region,finance_plan)%>%
  summarise(total = sum(abortion)) %>% 
  plot_ly(
    x = ~finance_plan, y = ~total, color = ~region, 
    type = "bar", colors = "viridis") %>% 
  layout(
    title = "Induced Abortions in NY State by Financial Plan Status", 
    xaxis = list(title = "Type of Financial Plan"),
    yaxis = list(title = "Abortions per 1,000 Live Births"))

finance_plot_nys
```

Highest rates of abortion are found among those did not report an insurance plan and thsoe who reported as having other insurance. The highest rate of abortion, when stratified by insurance enrollment, is found among those residing in the Capital/Saratoga region, and reporting as `Not Stated` insurance category with 168.13k induced abortions per 1000 live births. Lowest rates of abortion are found among those that are enrolled in Medicaid. There were some regions that are missing data in the plot because the data is missing for some counties in the original data set.
