---
title: "New York City Exploratory Analysis "
---

```{r, message = FALSE, echo = FALSE}
library(tidyverse)
library(janitor)
library(ggridges)
library(patchwork)
library(plotly)

knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
	)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
  )
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


## Five Boroughs of New York City

### Total

First, we want to look at the overall trends in abortion rates in the five boroughs of New York City (Manhattan, Bronx, Queens, Brooklyn, Staten Island) across the years 2014 - 2019.

```{r}
abortion_year = read_csv("data/abortion_year_borough.csv")

year_table_nyc = 
  abortion_year %>% 
  filter(borough == "New York City") %>% 
  mutate(borough = fct_reorder(borough, total)) %>% 
  select(year, total) %>% 
  knitr::kable(caption = "Abortion Rates in New York City (2014-2019)", col.names = c("Year", "Abortion rate per 1000 live births"), align = 'lc', format = "simple")

year_table_nyc

city_year = abortion_year %>% 
  filter(borough == "New York City") %>% 
  mutate(borough = fct_reorder(borough, total)) %>% 
  plot_ly(y = ~total, x = ~year, color = ~borough, type = "scatter", mode = "line", colors = "viridis")

city_year
```

From the data points and graph, it is evident that abortion rates are overall decreasing over the years, however, we see a small increase from 2018 to 2019. For our project, we will be analyzing data from 2019 which is the most recent dataset. 

```{r}
borough_year= 
  abortion_year %>% 
  filter(!borough == "New York State") %>%
  filter(!borough == "New York City") %>%
  mutate(borough = fct_reorder(borough, total)) %>% 
  select(year, borough, total) %>% 
  pivot_wider(
    names_from = "borough",
    values_from = "total") %>% 
  knitr::kable(caption = "Abortion Rates in New York City by Boroughs (2014-2019)", col.names = c("Year", "Bronx", "Brooklyn", "Manhattan", "Queens", "Staten Island"), align = 'lccccc', format = "simple")

borough_year

borough_year_plot = abortion_year %>% 
  filter(!borough == "New York State") %>%
  filter(!borough == "New York City") %>%
  mutate(borough = fct_reorder(borough, total)) %>% 
  plot_ly(y = ~total, x = ~year, color = ~borough, type = "scatter", mode = "line", colors = "viridis")

borough_year_plot
```


```{r}
data = read_csv(file = "./data/master.csv") %>%
  janitor::clean_names()
```

#### Age

```{r}
age_nyc = 
  data %>%
  select(1:7) %>%
  slice(3:7)

age_nyc %>%
  pivot_longer(
    age_less_20:age_plus_40,
    names_to = "age", 
    values_to = "abortion",
    names_prefix = "age_"
  ) %>%
  mutate(
    age = replace(age, age == "less_20", "<20"),
    age = replace(age, age == "20_24", "20-24"),
    age = replace(age, age == "25_29", "25-29"),
    age = replace(age, age == "30_34", "30-34"),
    age = replace(age, age == "35_39", "35-39"),
    age = replace(age, age == "plus_40", "40+"),
    age = factor(age)
  )%>% 
  plot_ly(x = ~age, y = ~abortion, color = ~borough, type = "bar", colors = "viridis") %>%   
  layout(
    title = 'Induced Abortions in NYC by Age', 
    xaxis = list(title = "Age (Years)"),
    yaxis = list(title = 'Abortions per 1,000 Live Births'))

```

From the bar plots, it is evident that those who age less than 20 have the highest abortion ratios across the boroughs in NYC in 2019. These ratios eem to decrease as age categories increase. When comparing between boroughs, Manhattan had the highest abortion ratios in categories `<20` and `25-29`. Staten Island had the lowest abortion ratios overall, except for category `20-24`. The highest abortion rate across all boroughs and age categories was for Manhattan for `< 20` with 1964.9 induced abortions per 1,000 live births. The lowest abortion ratio across all boroughs and age categories was in Staten Island for `35-39` with 130 induced abortions per 1,000 live births. 

#### Race

```{r}
race_nyc = 
  data %>%
  select(1,9:12) %>%
  slice(3:7)

race_nyc %>%
  pivot_longer(
    nh_white_only_ratio:h_total,
    names_to = "race", 
    values_to = "abortion"
  ) %>%
    mutate(
    race = replace(race, race == "nh_white_only_ratio", "Non-Hispanic White Only"),
    race = replace(race, race == "nh_black_only_ratio", "Non-Hispanic Black Only"),
    race = replace(race, race == "nh_other_ratio", "Non-Hispanic Other"),
    race = replace(race, race == "h_total", "Hispanic"),
    race = factor(race)
    )  %>% 
  plot_ly(x = ~race, y = ~abortion, color = ~borough, type = "bar", colors = "viridis") %>%   
  layout(
    title = "Induced Abortions in NYC by Race/Ethnicity", 
    xaxis = list(title = "Race"),
    yaxis = list(title = "Abortions per 1,000 Live Births"))

```

From the bar plots, it is evident that those who identified as Non-Hispanic Black Only had the highest abortion ratios across all boroughs in NYC. Among those who identified as Non-Hispanic Black Only, rates were highest in Manhattan with 1,228.3 induced abortions per 1,000 live births. Those who identified as Non-Hispanic White and Non-Hispanic Other had some of the lowest abortion ratios in NYC. The lowest rates across all boroughs and race groups was for those who identified as Non-Hispanic White only and lived in Brooklyn with a rates of 88.6 induced abortions per 1,000 live births. 

#### Financial Plans

```{r}
finance_plan_nyc= 
  data %>%
  select(1,15:18) %>%
  slice(3:7)

finance_plan_nyc %>%  
  pivot_longer(
    medicaid:not_stated,
    names_to = "finance_plan", 
    values_to = "abortion"
  ) %>%
 mutate(
    finance_plan = replace(finance_plan, finance_plan == "medicaid", "Medicaid"),
    finance_plan = replace(finance_plan, finance_plan == "self_pay", "Self Pay"),
    finance_plan = replace(finance_plan, finance_plan == "other_insurance", "Other Insurance"),
    finance_plan = replace(finance_plan, finance_plan == "not_stated", "Not Stated"),
    finance_plan = factor(finance_plan)
    )  %>% 
  plot_ly(x = ~finance_plan, y = ~abortion, color = ~borough, type = "bar", colors = "viridis") %>%   
  layout(
    title = "Induced Abortions in NYC by Financial Plan Status", 
    xaxis = list(title = "Type of Financial Plan"),
    yaxis = list(title = "Abortions per 1,000 Live Births"))
```

From the bar plots, those who were categorized as having Other Insurance had the highest abortion rates across all financial plans in NYC with a rate of 62.86k induced abortions per 1000 live births. When calculating abortion rates for financial plans, we manually calculated these rates from (induced abortions/ live births)*1000 as they were not provided like in other data sets. A huge flaw in these numbers is that those who have induced abortions and those who give births tend to use different insurance plans. People who give birth are unlikely to use `Other Insurance` in comparison to those who have induced abortions due to the inherent differences in procedures. Those who used Self Pay for induced abortions had a relatively high abortion rates as well. This would make sense as well given the nature of the procedure and possible restrictions by the health insurance plan to cover abortions. Medicaid had the lowest abortion rates across all financial plans, and the rates were lowest with 287 induced abortions per 1000 live births.
