---
title: "NY State Statistical Analysis"
output: 
  html_document: 
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

&nbsp; 

```{r setup, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r data, message=FALSE, include=FALSE}
abortion_data = read_csv("data/master.csv")
```

## Abortion Trends in NY State

&nbsp; 

### Proportion Test - NY State Regions
***

**Do overall abortion rates differ across NY state regions?**

**$H_0$**: The abortion rate per 1000 live births is the same across all NY state regions.

**$H_1$**: The abortion rate per 1000 live births is not the same across all NY state regions.

```{r data_NY, message=FALSE, include=FALSE}
NY_year = read_csv("data/NY_year.csv")
```

```{r tidy_NY}
# Tidy data
NY_totals =
  NY_year %>%
  select(borough, total) %>%
  slice(1:10) %>%
  mutate(
    n = total,
    total = 1000)

NY_totals %>% 
  select(-total) %>%
  knitr::kable(caption = "Proportion T-Test for NY across all Regions",
  col.names = c("Region", "Abortions per 1000 live births"), align = c('l', 'c'), format = "simple")
```

```{r prop_NY}
# Run 10 prop t-test
prop.test(NY_totals$n, NY_totals$total,
          p = NULL, alternative = "two.sided")
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates per 1000 live births differs across NY state regions.

Because this suggests that abortion rates differs across NY state regions, we now can assess whether or not this holds true across counties.

&nbsp; 

### Proportion Test - NY State Counties
***

**Do overall abortion rates differ across NY state counties?**

**$H_0$**: The abortion rates per 1000 live births is the same across all NY state counties.

**$H_1$**: The abortion rates per 1000 live births is not the same across all NY state counties.

```{r tidy_NY_counties}
# Tidy data
NY_totals_counties =
  abortion_data %>%
  select(borough, total) %>%
  slice(-(1:2)) %>%
  mutate(
    n = total,
    total = 1000)

NY_totals_counties %>% 
  select(-total) %>%
  head(10) %>%
  knitr::kable(caption = "Proportion T-Test for NY across all Counties",
  col.names = c("County", "Abortions per 1000 live births"), align = c('l', 'c'), format = "simple")
```

Only the first 10 counties are printed.

```{r prop_NY_counties}
# Run 58 prop t-test
prop.test(NY_totals_counties$n, NY_totals_counties$total,
          p = NULL, alternative = "two.sided")
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates per 1000 live births differs across NY state counties.

&nbsp; 

### Chi Square Test - Race and NY State Counties
***

Due to missing data, a chi-squared analysis was not run for race and NY state regions.

**Do overall abortion rates differ across NY state counties by race?**

**$H_0$**: The expected abortion rates in each race category are the same across all NY state counties.

**$H_1$**: The expected abortion rates in each race category are not the same across all NY state counties.

```{r NY_tidy_race_counties}
# Tidy data
NY_race_clean = 
  abortion_data %>%
  slice(-(1:2)) %>%
  select(-(age_less_20:nh_total)) %>% select(-(h_white_only_ratio:region)) %>%
  drop_na() %>%
  mutate(
    hispanic = h_total*1000,
    nh_white_only = nh_white_only_ratio*1000,
    nh_black_only = nh_black_only_ratio*1000,
    nh_other = nh_other_ratio*1000
  ) %>%
  select(-(nh_white_only_ratio:h_total))

NY_race_clean %>% 
  head(10) %>%
  knitr::kable(caption = "Chi-Square Test for Race and Counties", col.names = c("County", "Hispanic", "Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Other"), align = c('l', 'c', 'c', 'c', 'c','c','c'), format = "simple")
```

Only the first 10 counties are printed.

```{r NY_chi-sq_race_counties}
NY_race_clean =
  NY_race_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(NY_race_clean)
```

```{r NY_x_crit_race_counties}
x_crit = qchisq(0.95, 135)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates in each race category across counties is different.

&nbsp; 

### Chi Square Test - Finance Plans and NY State Regions
***

**Do overall abortion rates differ across NY state regions by finance plan?**

**$H_0$**: The expected abortion rates in each finance plan category are the same across all NY state regions.

**$H_1$**: The expected abortion rates in each finance plan category are not the same across all NY state regions.

```{r data_NY_finance, include=FALSE, message=FALSE}
NY_finance = read_csv("data/NY_finance.csv")
```

```{r NY_tidy_finance}
# Tidy data
NY_finance_clean = 
  NY_finance %>%
  mutate(
    medicaid = medicaid*1000,
    self_pay = self_pay*1000,
    other_insurance = other_insurance*1000,
    not_stated = not_stated*1000
  ) %>%
  select(-total)

NY_finance_clean %>%
  knitr::kable(caption = "Chi-Square Test Finance Plan and Regions", col.names = c("Region", "Medicaid", "Self-Pay", "Other Insurance", "Not Stated"), align = c('l', 'c', 'c', 'c', 'c'), format = "simple" )
```

```{r NY_chi-sq_finance}
NY_finance_clean =
  NY_finance_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(NY_finance_clean)
```

```{r NY_x_crit_finance}
x_crit = qchisq(0.95, 27)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates in each finance category across regions is different.

&nbsp; 

### Chi Square Test - Finance Plan and NY State Counties
***

**Do overall abortion rates differ across NY state counties by finance plan?**

**$H_0$**: The expected abortion rates in each finance plan category are the same across all NY state counties.

**$H_1$**: The expected abortion rates in each finance plan category are not the same across all NY state counties.

```{r NY_finance_counties}
# Tidy data
NY_finance_counties = 
  abortion_data %>%
  slice(-(1:2)) %>%
  select(-(age_less_20:nh_total)) %>% select(-(h_white_only_ratio:region)) %>%
  drop_na() %>%
  mutate(
    hispanic = h_total*1000,
    nh_white_only = nh_white_only_ratio*1000,
    nh_black_only = nh_black_only_ratio*1000,
    nh_other = nh_other_ratio*1000
  ) %>%
  select(-(nh_white_only_ratio:h_total))

NY_finance_counties %>% 
  head(10) %>%
  knitr::kable(caption = "Chi-Square Test for Race and Counties", col.names = c("County", "Hispanic", "Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Other"), align = c('l', 'c', 'c', 'c', 'c'), format = "simple")
```

Only the first 10 counties are printed.

```{r NY_chi-sq_finance_counties}
NY_finance_counties =
  NY_finance_counties %>%
  select(-borough)

# Run chi-square test
chisq.test(NY_finance_counties)
```

```{r NY_x_crit_finance_counties}
x_crit = qchisq(0.95, 135)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates in each finance category across counties is different.

&nbsp; 

### Chi Square Test - Age and NY State Counties
***

Due to missing data, a chi-squared analysis was not run for race and NY state regions.

**Do overall abortion rates differ across NY state counties by age?**

**$H_0$**: The expected abortion rates in each age category are the same across all NY state counties.

**$H_1$**: The expected abortion rates in each age category are not the same across all NY state counties.

```{r NY_tidy_age_counties}
# Tidy data
NY_age_clean = 
  abortion_data %>%
  slice(-(1:2)) %>%
  select(-(nh_total:region)) %>%
  drop_na() %>%
  mutate(
    age_less_20 = age_less_20*1000,
    age_20_24 = age_20_24*1000,
    age_25_29 = age_25_29*1000,
    age_30_34 = age_30_34*1000,
    age_35_39 = age_35_39*1000,
    age_plus_40 = age_plus_40*1000
  )

NY_age_clean %>% 
  head(10) %>%
  knitr::kable(caption = "Chi-Square Test for Age and Counties", col.names = c("County", "< 20 years", "20-24 years", "25-29 years", "30-34 years", "35-39 years", "40+ years"), align = c('l', 'c', 'c', 'c', 'c', 'c', 'c'), format = "simple" )
```

Only the first 10 counties are printed.

```{r NY_chi-sq_age_counties}
NY_age_clean =
  NY_age_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(NY_age_clean)
```

```{r NY_x_crit_age}
x_crit = qchisq(0.95, 275)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates in each age category across counties is different.

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div> 