---
title: "Statistical Analysis of Abortion Trends Among NYC and Compared to NY State"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

```{r setup, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r data, message=FALSE, include=FALSE}
abortion_data = read_csv("data/master.csv")
```

## Abortion Trends in NYC

### *Proportion T-Test - Boroughs*

**Do overall abortion rates differ across boroughs?**

**$H_0$**: The abortion rate per 1000 live births is the same across all boroughs.

**$H_1$**: The abortion rate per 1000 live births is not the same across all boroughs.

```{r tidy_borough}
# Tidy data
abortion_totals_borough =
  abortion_data %>%
  select(borough, total) %>%
  slice(3:7) %>%
  mutate(
    n = total,
    total = 1000)

abortion_totals_borough %>%
  select (-total) %>%
  knitr::kable(caption = "Proportion T-Test for NYC across all Boroughs", col.names = c("Borough", "Abortion Rate per 1000 live births"), align = c('l', 'c'), format = "simple")
```

```{r prop_boroughs}
# Run 5 prop t-test
prop.test(abortion_totals_borough$n, abortion_totals_borough$total,
          p = NULL, alternative = "two.sided")
```

**Inference:**

- From the above results, the p-value is **less than the 5% level of significance**. Therefore, we **reject the null hypothesis**. 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates per 1000 births live differs across boroughs.

### *Chi Square Test - Race and Boroughs*

**Do overall abortion rates differ across boroughs by race?**

**$H_0$**: The expected abortion rate in each race category are the same across all boroughs.

**$H_1$**: The expected abortion rate in each race category are not the same across all boroughs.

```{r tidy_race}
# Tidy data
abortion_race_clean = 
  abortion_data %>%
  slice(3:7) %>%
  select(-(age_less_20:nh_total)) %>% select(-(h_white_only_ratio:region)) %>%
  mutate(
    hispanic = h_total*1000,
    nh_white_only = nh_white_only_ratio*1000,
    nh_black_only = nh_black_only_ratio*1000,
    nh_other = nh_other_ratio*1000
  ) %>%
  select(-(nh_white_only_ratio:h_total))

abortion_race_clean %>% 
  knitr::kable(caption = "Chi-Square Test for Race and Boroughs", col.names = c("Borough", "Hispanic", "Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Other"), align = c('l', 'c', 'c', 'c', 'c'), format = "simple")
```

```{r chi-sq_race}
abortion_race_clean =
  abortion_race_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(abortion_race_clean)
```

```{r x_crit_race}
x_crit = qchisq(0.95, 12)
x_crit
```

**Inference:** 

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, we **reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that the abortion rate in each race category across boroughs is different. 

### *Chi Square Test - Finance Plan and Boroughs*

**Do overall abortion rates differ across boroughs by finance plan?**

**$H_0$**: The expected abortion rates in each finance plan category are the same across all boroughs.

**$H_1$**: The expected abortions rates in each finance plan category are not the same across all boroughs.

```{r tidy_finance}
# Tidy data
abortion_finance_clean = 
  abortion_data %>%
  slice(3:7) %>%
  select(-(age_less_20:h_black_only_ratio)) %>%
  mutate(
    medicaid = medicaid*1000,
    self_pay = self_pay*1000,
    other_insurance = other_insurance*1000,
    not_stated = not_stated*1000
  ) %>%
  select(-(total:region))

abortion_finance_clean %>% 
  knitr::kable(caption = "Chi-Square Test for Finance Plan and Boroughs", col.names = c("Borough", "Medicaid", "Self-Pay", "Other Insurance", "Not Stated"), align = c('l', 'c', 'c', 'c', 'c'), format = "simple" )
```

```{r chi-squ_finance}
abortion_finance_clean =
  abortion_finance_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(abortion_finance_clean)
```

```{r x_crit_finance}
x_crit = qchisq(0.95, 12)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that the abortion rates in each finance category across borough is different.

### *Chi Square Test - Age and Boroughs*

**Do overall abortion rates differ across boroughs by age?**

**$H_0$**: The expected abortion rates in each age category are the same across all boroughs.

**$H_1$**: The expected abortion rates in each age category are not the same across all boroughs.

```{r tidy_age}
# Tidy data
abortion_age_clean = 
  abortion_data %>%
  slice(3:7) %>%
  select(-(nh_total:region)) %>%
  mutate(
    age_less_20 = age_less_20*1000,
    age_20_24 = age_20_24*1000,
    age_25_29 = age_25_29*1000,
    age_30_34 = age_30_34*1000,
    age_35_39 = age_35_39*1000,
    age_plus_40 = age_plus_40*1000
  )

abortion_age_clean %>% 
  knitr::kable(caption = "Chi-Square Test for Age and Boroughs", col.names = c("Borough", "< 20 years", "20-24 years", "25-29 years", "30-34 years", "35-39 years", "40+ years"), align = c('l', 'c', 'c', 'c', 'c', 'c', 'c'), format = "simple" )
```

```{r chi-sq_age}
abortion_age_clean =
  abortion_age_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(abortion_age_clean)
```

```{r x_crit_age}
x_crit = qchisq(0.95, 20)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that the overall abortion rate in each age category across borough is different.

## *Abortion Trends in NY State*

### *Proportion Test - NY State Regions*

**Do overall abortion rates differ across NY state regions?**

**$H_0$**: The abortion rate per 1000 live births is the same across all NY state regions.

**$H_1$**: The abortion rate per 1000 live births is not the same across all NY state regions.

```{r data_NY, message=FALSE, include=FALSE}
NY_year = read_csv("data/NY_year.csv")
```

```{r tidy_NY}
# Tidy data
NY_totals =
  NY_year %>%
  select(borough, total) %>%
  slice(1:10) %>%
  mutate(
    n = total,
    total = 1000)

NY_totals %>% 
  select(-total) %>%
  knitr::kable(caption = "Proportion T-Test for NY across all Regions",
  col.names = c("Region", "Abortions per 1000 live births"), align = c('l', 'c'), format = "simple")
```

```{r prop_NY}
# Run 10 prop t-test
prop.test(NY_totals$n, NY_totals$total,
          p = NULL, alternative = "two.sided")
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates per 1000 live births differs across NY state regions.

Because this suggests that abortion rates differs across NY state regions, we now can assess whether or not this holds true across counties.

### *Proportion Test - NY State Counties*

**Do overall abortion rates differ across NY state counties?**

**$H_0$**: The abortion rates per 1000 live births is the same across all NY state counties.

**$H_1$**: The abortion rates per 1000 live births is not the same across all NY state counties.

```{r tidy_NY_counties}
# Tidy data
NY_totals_counties =
  abortion_data %>%
  select(borough, total) %>%
  slice(-(1:2)) %>%
  mutate(
    n = total,
    total = 1000)

NY_totals_counties %>% 
  select(-total) %>%
  head(10) %>%
  knitr::kable(caption = "Proportion T-Test for NY across all Counties",
  col.names = c("County", "Abortions per 1000 live births"), align = c('l', 'c'), format = "simple")
```

Only the first 10 counties are printed.

```{r prop_NY_counties}
# Run 58 prop t-test
prop.test(NY_totals_counties$n, NY_totals_counties$total,
          p = NULL, alternative = "two.sided")
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates per 1000 live births differs across NY state counties.

### *Chi Square Test - Race and NY State Counties*

Due to missing data, a chi-squared analysis was not run for race and NY state regions.

**Do overall abortion rates differ across NY state counties by race?**

**$H_0$**: The expected abortion rates in each race category are the same across all NY state counties.

**$H_1$**: The expected abortion rates in each race category are not the same across all NY state counties.

```{r NY_tidy_race_counties}
# Tidy data
NY_race_clean = 
  abortion_data %>%
  slice(-(1:2)) %>%
  select(-(age_less_20:nh_total)) %>% select(-(h_white_only_ratio:region)) %>%
  drop_na() %>%
  mutate(
    hispanic = h_total*1000,
    nh_white_only = nh_white_only_ratio*1000,
    nh_black_only = nh_black_only_ratio*1000,
    nh_other = nh_other_ratio*1000
  ) %>%
  select(-(nh_white_only_ratio:h_total))

NY_race_clean %>% 
  head(10) %>%
  knitr::kable(caption = "Chi-Square Test for Race and Counties", col.names = c("County", "Hispanic", "Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Other"), align = c('l', 'c', 'c', 'c', 'c','c','c'), format = "simple")
```

Only the first 10 counties are printed.

```{r NY_chi-sq_race_counties}
NY_race_clean =
  NY_race_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(NY_race_clean)
```

```{r NY_x_crit_race_counties}
x_crit = qchisq(0.95, 135)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates in each race category across counties is different.

### *Chi Square Test - Finance Plans and NY State Regions*

**Do overall abortion rates differ across NY state regions by finance plan?**

**$H_0$**: The expected abortion rates in each finance plan category are the same across all NY state regions.

**$H_1$**: The expected abortion rates in each finance plan category are not the same across all NY state regions.

```{r data_NY_finance, include=FALSE, message=FALSE}
NY_finance = read_csv("data/NY_finance.csv")
```

```{r NY_tidy_finance}
# Tidy data
NY_finance_clean = 
  NY_finance %>%
  mutate(
    medicaid = medicaid*1000,
    self_pay = self_pay*1000,
    other_insurance = other_insurance*1000,
    not_stated = not_stated*1000
  ) %>%
  select(-total)

NY_finance_clean %>%
  knitr::kable(caption = "Chi-Square Test Finance Plan and Regions", col.names = c("Region", "Medicaid", "Self-Pay", "Other Insurance", "Not Stated"), align = c('l', 'c', 'c', 'c', 'c'), format = "simple" )
```

```{r NY_chi-sq_finance}
NY_finance_clean =
  NY_finance_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(NY_finance_clean)
```

```{r NY_x_crit_finance}
x_crit = qchisq(0.95, 27)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates in each finance category across regions is different.

### *Chi Square Test - Finance Plan and NY State Counties*

**Do overall abortion rates differ across NY state counties by finance plan?**

**$H_0$**: The expected abortion rates in each finance plan category are the same across all NY state counties.

**$H_1$**: The expected abortion rates in each finance plan category are not the same across all NY state counties.

```{r NY_finance_counties}
# Tidy data
NY_finance_counties = 
  abortion_data %>%
  slice(-(1:2)) %>%
  select(-(age_less_20:nh_total)) %>% select(-(h_white_only_ratio:region)) %>%
  drop_na() %>%
  mutate(
    hispanic = h_total*1000,
    nh_white_only = nh_white_only_ratio*1000,
    nh_black_only = nh_black_only_ratio*1000,
    nh_other = nh_other_ratio*1000
  ) %>%
  select(-(nh_white_only_ratio:h_total))

NY_finance_counties %>% 
  head(10) %>%
  knitr::kable(caption = "Chi-Square Test for Race and Counties", col.names = c("County", "Hispanic", "Non-Hispanic White", "Non-Hispanic Black", "Non-Hispanic Other"), align = c('l', 'c', 'c', 'c', 'c'), format = "simple")
```

Only the first 10 counties are printed.

```{r NY_chi-sq_finance_counties}
NY_finance_counties =
  NY_finance_counties %>%
  select(-borough)

# Run chi-square test
chisq.test(NY_finance_counties)
```

```{r NY_x_crit_finance_counties}
x_crit = qchisq(0.95, 135)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates in each finance category across counties is different.

### *Chi Square Test - Age and NY State Counties*

Due to missing data, a chi-squared analysis was not run for race and NY state regions.

**Do overall abortion rates differ across NY state regions by counties?**

**$H_0$**: The expected abortion rates in each age category are the same across all NY state counties.

**$H_1$**: The expected abortion rates in each age category are not the same across all NY state counties.

```{r NY_tidy_age_counties}
# Tidy data
NY_age_clean = 
  abortion_data %>%
  slice(-(1:2)) %>%
  select(-(nh_total:region)) %>%
  drop_na() %>%
  mutate(
    age_less_20 = age_less_20*1000,
    age_20_24 = age_20_24*1000,
    age_25_29 = age_25_29*1000,
    age_30_34 = age_30_34*1000,
    age_35_39 = age_35_39*1000,
    age_plus_40 = age_plus_40*1000
  )

NY_age_clean %>% 
  head(10) %>%
  knitr::kable(caption = "Chi-Square Test for Age and Counties", col.names = c("County", "< 20 years", "20-24 years", "25-29 years", "30-34 years", "35-39 years", "40+ years"), align = c('l', 'c', 'c', 'c', 'c', 'c', 'c'), format = "simple" )
```

Only the first 10 counties are printed.

```{r NY_chi-sq_age_counties}
NY_age_clean =
  NY_age_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(NY_age_clean)
```

```{r NY_x_crit_age}
x_crit = qchisq(0.95, 275)
x_crit
```

**Inference:**

- From the above results, **$X^2$ is greater than the test-statistic at the 5% level of significance**. Therefore, **we reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that overall abortion rates in each age category across counties is different.

## Abortions Rates Comparing NYC and NY State

### *Proportion T-test*

**Do overall abortion rates differ between NYC and NY state?**

**$H_0$**: The abortion rate per 1000 births is the same between NYC and NY state.

**$H_1$**: The abortion rate per 1000 births differs between NYC and NY state.

```{r tidy_totals}
# Tidy data
abortion_totals =
  abortion_data %>%
  select(borough, total) %>%
  head(2) %>%
  mutate(
    n = total,
    total = 1000)

abortion_totals %>% 
  select (-total) %>%
  knitr::kable(caption = "Proportion T-Test for NYC and NY State", col.names = c("Region", "Abortion Rate per 1000 births"), align = c('l', 'c'), format = "simple")
```

```{r prop}
# Run prop t-test
prop.test(abortion_totals$n, abortion_totals$total,
          p = NULL, alternative = "two.sided")
```

**Inference:**

- From the above results, the p-value is **less than the 5% level of significance.** Therefore, we **reject the null hypothesis.** 

- The data provides sufficient evidence to support the hypothesis that the abortion rate per 1000 births differs between NYC and NY state.