---
title: "Statistical Analysis of Abortion Trends Among NYC and Compared to NY State"
---

```{r setup, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r data, message=FALSE, include=FALSE}
abortion_data = read_csv("data/abortion_data.csv")
abortion_race = read_csv("data/abortion_race.csv")
abortion_finance = read_csv("data/abortion_finance.csv")
```

## *Abortions Rates Comparing NYC and NY State*

*Proportion T-test*

Do overall abortion rates differ between NYC and NY state?

H0: The number of abortions per 1000 births is the same between NYC and NY state.

H1: The number of abortions per 1000 births differs between NYC and NY state.

```{r tidy_totals}
# Tidy data
abortion_totals =
  abortion_data %>%
  select(borough, total) %>%
  head(2) %>%
  mutate(
    n = total,
    total = 1000)

abortion_totals %>% 
  knitr::kable(caption = "Results Table")
```

```{r prop}
# Run prop t-test
prop.test(abortion_totals$n, abortion_totals$total,
          p = NULL, alternative = "two.sided")
```

From the above results, the p-value is less than the 5% level of significance. Therefore, we reject the null hypothesis. The data provides sufficient evidence to support the hypothesis that the number of abortions per 1000 births differs between NYC and NY state.

## *Abortion Trends in NYC*

*Proportion Test*

Do overall abortion rates differ across boroughs?

H0: The number of abortions per 1000 births is the same across all boroughs.

H1: The number of abortions per 1000 births is not the same across all boroughs.

```{r tidy_borough}
# Tidy data
abortion_totals_borough =
  abortion_data %>%
  select(borough, total) %>%
  slice(3:7) %>%
  mutate(
    n = total,
    total = 1000)

abortion_totals_borough %>% 
  knitr::kable(caption = "Results Table")
```

```{r prop_boroughs}
# Run 5 prop t-test
prop.test(abortion_totals_borough$n, abortion_totals_borough$total,
          p = NULL, alternative = "two.sided")
```

From the above results, the p-value is less than the 5% level of significance. Therefore, we reject the null hypothesis. The data provides sufficient evidence to support the hypothesis that the number of abortions per 1000 births differs across boroughs.

*Chi Square Test - Race and Boroughs*

Do overall abortion rates differ across boroughs, controlling for race?

H0: The number of abortions per 1000 births is the same across all boroughs, controlling for race.

H1: The number of abortions per 1000 births is not the same across all boroughs, controlling for race.

```{r tidy_race}
# Tidy data
abortion_race_clean = 
  abortion_race %>%
  slice(3:7) %>%
  mutate(
    hispanic = hispanic_ratio*1000,
    h_white_only = h_white_only_ratio*1000,
    h_black_only = h_black_only_ratio*1000,
    non_hispanic = non_hispanic_ratio*1000,
    nh_white_only = nh_white_only_ratio*1000,
    nh_black_only = nh_black_only_ratio*1000,
    nh_other = nh_other_ratio*1000
  ) %>%
  select(-(hispanic_ratio:nh_other_ratio)) %>% select(-total)

abortion_race_clean %>% 
  knitr::kable(caption = "Results Table")
```

```{r chi-sq}
abortion_race_clean =
  abortion_race_clean %>%
  select(-borough)

# Run chi-square test
chisq.test(abortion_race_clean)
```

```{r x_crit}
x_crit = qchisq(0.95, 24)
x_crit
```

From the above results, X^2 is greater than the test-statistic at the 5% level of significance. Therefore, we reject the null hypothesis. The data provides sufficient evidence to support the hypothesis that the number of abortions 1000 births is not the same across all boroughs, controlling for race. 

