---
title: "Map"
date: "2022-12-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
library(patchwork)
library(readxl)
library(rvest)
library(httr)
library(dplyr)
library(leaflet)

library(tigris)
library(sp)
library(ggmap)
library(maptools)
library(broom)
library(rgdal)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

## Exploratory Analysis

Get the data

```{r data_import}
abortion_final = read_csv(file = "data/abortion_data.csv") %>%
    janitor::clean_names()
```

```{r}
library(rgeos)
library(maptools)
library(geojsonio)
library(ggplot2)

# this is the geojson of the NYC community districts
URL <- "http://services5.arcgis.com/GfwWNkhOj9bNBqoJ/arcgis/rest/services/nycd/FeatureServer/0/query?where=1=1&outFields=*&outSR=4326&f=geojson"
fil <- "nyc_community_districts.geojson"
if (!file.exists(fil)) download.file(URL, fil)

nyc_districts <- geojson_read(fil, what="sp")

# the @data slot of nyc_districts has 2 fields. BoroCD is the district #
nyc_districts_map <- fortify(nyc_districts, region="BoroCD")

# let's see which id is what
mids <- cbind.data.frame(as.data.frame(gCentroid(nyc_districts, byid=TRUE)), 
                         id=nyc_districts$BoroCD)

gg <- ggplot()
gg <- gg + geom_map(data=nyc_districts_map, map=nyc_districts_map,
                    aes(x=long, y=lat, map_id=id),
                    color="#2b2b2b", size=0.15, fill=NA)
gg <- gg + geom_text(data=mids, aes(x=x, y=y, label=id), size=2)
gg <- gg + coord_map()
gg <- gg + ggthemes::theme_map()
gg

```

```{r}
leaflet() %>%
  addTiles() %>%
  setView(-74.00, 40.71, zoom = 10) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addMarkers(lat = 40.71, lng = -74, popup = "Manhattan") %>%
  addMarkers(lat = 40.83, lng = -73.87, popup = "Bronx") %>%
  addMarkers(lat = 40.73, lng = -73.79, popup = "Queen") %>%
  addMarkers(lat = 40.58, lng = -74.15, popup = "Staten Island") %>%
  addMarkers(lat = 40.68, lng = -73.94, popup = "Brooklyn")

nyc_tracts <- county_subdivisions(state = '36', county = c('061','047','081','005','085'))

summary(nyc_tracts)
plot(nyc_tracts)


```


## `Plotly`

Trying out the built-in state geometries within the `plotly` package.


```{r}
library(tidyverse)
library(plotly)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r}
abortion_final = read_csv(file = "data/abortion_data.csv") %>%
    janitor::clean_names()
```

```{r}
l = list(colour = toRGB("white"), width = 2)

g = list(scope = "usa")

test = abortion_final %>% 
  select(borough, total) %>% 
  plot_geo(locationmode = "USA-states", locations = "NY") %>%
  add_trace(
    z = ~total, locations = ~borough,
    color = ~total, colors = "viridis"
  ) %>% 
  colorbar(title = "") %>% 
  layout(geo = g)
```


Couldn't get display of state-only level data. Trying out the `tidycensus` package next...

## `tidycensus`

Working with non-census data...

```{r}
library(tigris)
```

```{r}
ny_counties = 
  counties(state = "36", cb = TRUE, resolution = "20m") 

write_csv(ny_counties, file = "data/ny_counties_geo.csv")
```

This method seems promising, using methods described in the `tidycensus` [website](https://walker-data.com/census-r/mapping-census-data-with-r.html#cartographic-workflows-with-non-census-data), chapter 6.4. Will wait for a merged dataset with geographic and abortion information to try make an interactive map.


```{r}
merged_data = read_csv("data/master.csv")
```

