---
title: "NY_race"
author: "Maisie Sun"
date: "2022-12-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(data.table)
library(janitor)
library(ggridges)
library(patchwork)
library(readxl)
library("writexl")
library(rvest)
library(httr)
library(dplyr)
library(plotly)
knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

## Financial Plans

Get the data

```{r}
url = "https://www.health.ny.gov/statistics/vital_statistics/2019/table23.htm"
url1 = "https://www.health.ny.gov/statistics/vital_statistics/2019/table09.htm"

induced_abortion_race = 
  read_html(url) %>%
  html_table(header = FALSE) %>%
  first() %>%
  janitor::clean_names()
live_birth_race = 
  read_html(url1) %>%
  html_table(header = FALSE) %>%
  first() %>%
  janitor::clean_names()
```

data cleaning NH

```{r data_cleaning_nhrace}
clean_nh = 
  induced_abortion_race %>%
  select(1,2,4,6,8,10) %>% 
  purrr::set_names(c("borough", "total", "Non_Hispanic_Count", "NH_White_Only_Count", "NH_Black_Only_Count", "NH_Other_Count")) %>% 
  slice(7:70) %>% 
  filter(!row_number() %in% c(6, 7)) %>%
  mutate(
    borough = str_replace(borough, "Kings", "Brooklyn"),
    borough = ifelse(as.character(borough) == "New York", "Manhattan", as.character(borough)),
    borough = str_replace(borough, "Richmond", "Staten Island"),
    total = str_replace(total, ",", ""),
    Non_Hispanic_Count = str_replace(Non_Hispanic_Count, ",", ""),
    NH_White_Only_Count = str_replace(NH_White_Only_Count, ",", ""),
    NH_Black_Only_Count = str_replace(NH_Black_Only_Count, ",", ""),
    NH_Other_Count = str_replace(NH_Other_Count, ",", "")
    ) %>%
  mutate_at(c("total", "Non_Hispanic_Count", "NH_White_Only_Count", "NH_Black_Only_Count", "NH_Other_Count"), as.numeric)

```

data cleaning H

```{r data_cleaning_hrace}
clean_h = 
  induced_abortion_race %>%
  select(1,2,12,14,16) %>% 
  purrr::set_names(c("borough", "total", "Hispanic_Count", "H_White_Only_Count", "H_Black_Only_Count")) %>% 
  slice(7:70) %>% 
  filter(!row_number() %in% c(6, 7)) %>%
  mutate(
    borough = str_replace(borough, "Kings", "Brooklyn"),
    borough = ifelse(as.character(borough) == "New York", "Manhattan", as.character(borough)),
    borough = str_replace(borough, "Richmond", "Staten Island"),
    total = str_replace(total, ",", ""),
    Hispanic_Count = str_replace(Hispanic_Count, ",", ""),
    H_White_Only_Count = str_replace(H_White_Only_Count, ",", ""),
    H_Black_Only_Count = str_replace(H_Black_Only_Count, ",", "")
    ) %>%
  mutate_at(c("total", "Hispanic_Count", "H_White_Only_Count", "H_Black_Only_Count"), as.numeric)
```

data cleaning lb NH

```{r data_cleaning_nhrace}
clean_lb_nh = 
  live_birth_race %>%
  select(1:6) %>% 
  purrr::set_names(c("borough", "total", "Non_Hispanic_Count", "NH_White_Only_Count", "NH_Black_Only_Count", "NH_Other_Count")) %>% 
  slice(7:70) %>% 
  filter(!row_number() %in% c(6, 7)) %>%
  mutate(
    borough = str_replace(borough, "Kings", "Brooklyn"),
    borough = ifelse(as.character(borough) == "New York", "Manhattan", as.character(borough)),
    borough = str_replace(borough, "Richmond", "Staten Island"),
    total = str_replace(total, ",", ""),
    Non_Hispanic_Count = str_replace(Non_Hispanic_Count, ",", ""),
    NH_White_Only_Count = str_replace(NH_White_Only_Count, ",", ""),
    NH_Black_Only_Count = str_replace(NH_Black_Only_Count, ",", ""),
    NH_Other_Count = str_replace(NH_Other_Count, ",", "")
    ) %>%
  mutate_at(c("total", "Non_Hispanic_Count", "NH_White_Only_Count", "NH_Black_Only_Count", "NH_Other_Count"), as.numeric)

```

data cleaning lb H

```{r data_cleaning_hrace}
clean_lb_h = 
  live_birth_race %>%
  select(1,2,7:9) %>% 
  purrr::set_names(c("borough", "total", "Hispanic_Count", "H_White_Only_Count", "H_Black_Only_Count")) %>% 
  slice(7:70) %>% 
  filter(!row_number() %in% c(6, 7)) %>%
  mutate(
    borough = str_replace(borough, "Kings", "Brooklyn"),
    borough = ifelse(as.character(borough) == "New York", "Manhattan", as.character(borough)),
    borough = str_replace(borough, "Richmond", "Staten Island"),
    total = str_replace(total, ",", ""),
    Hispanic_Count = str_replace(Hispanic_Count, ",", ""),
    H_White_Only_Count = str_replace(H_White_Only_Count, ",", ""),
    H_Black_Only_Count = str_replace(H_Black_Only_Count, ",", "")
    ) %>%
  mutate_at(c("total", "Hispanic_Count", "H_White_Only_Count", "H_Black_Only_Count"), as.numeric)
```

#### Renaming

created regions based on this website: https://statejobs.ny.gov/assets/help/regionMapText.cfm

renaming abortion H data of counties into regions.

```{r rename abortion H}
rename_ia_h= clean_h %>% 
  transform(borough = gsub(pattern = "Albany|Columbia|Fulton|Greene|Montgomery|Rensselaer|Saratoga|Schenectady|Schoharie|Warren|Washington", replacement = "Saratoga", borough)) %>%
  transform(borough = gsub(pattern = "Franklin|Clinton|Essex|Hamilton|Essex/Hamilton|Hamilton/Essex", replacement = "Eastern Adirondacks", borough)) %>%
  transform(borough = gsub(pattern = "Herkimer|Jefferson|Lewis|Oneida|Oswego|St Lawrence", replacement = "Western Adirondacks", borough)) %>%
  transform(borough = gsub(pattern = "Broome|Cayuga|Chenango|Cortland|Madison|Onondaga|Otsego|Tioga|Tompkins", replacement = "Central New York", borough)) %>% 
  transform(borough=gsub(pattern="Chemung|Genesee|Livingston|Monroe|Ontario|Schuyler|Seneca|Steuben|Wayne|Yates",replacement="Finger Lakes", borough)) %>% 
  transform(borough=gsub(pattern="Allegany|Cattaraugus|Chautauqua|Erie|Niagara|Orleans|Wyoming",replacement="Western New York", borough)) %>% 
  transform(borough=gsub(pattern="Delaware|Dutchess|Orange|Putnam|Sullivan|Ulster",replacement="Hudson Valley", borough)) %>% 
  transform(borough=gsub(pattern="Rockland|Westchester", replacement="Westchester/Rockland", borough)) %>% 
  transform(borough=gsub(pattern="Suffolk|Nassau", replacement="Long Island", borough)) %>% 
    transform(borough=gsub(pattern="Bronx|Brooklyn|Manhattan|Queens|Staten Island", replacement="New York City", borough)) %>% 
  group_by(borough) %>% 
  summarize (
    total_rate= sum(total), 
    Hispanic_Count= sum(Hispanic_Count),
    H_White_Only_Count= sum(H_White_Only_Count), 
    H_Black_Only_Count= sum(H_Black_Only_Count))
```

renaming abortion NH data of counties into regions.

```{r rename abortion NH}
rename_ia_nh= clean_nh %>% 
  transform(borough = gsub(pattern = "Albany|Columbia|Fulton|Greene|Montgomery|Rensselaer|Saratoga|Schenectady|Schoharie|Warren|Washington", replacement = "Saratoga", borough)) %>%
  transform(borough = gsub(pattern = "Franklin|Clinton|Essex|Hamilton|Essex/Hamilton|Hamilton/Essex", replacement = "Eastern Adirondacks", borough)) %>%
  transform(borough = gsub(pattern = "Herkimer|Jefferson|Lewis|Oneida|Oswego|St Lawrence", replacement = "Western Adirondacks", borough)) %>%
  transform(borough = gsub(pattern = "Broome|Cayuga|Chenango|Cortland|Madison|Onondaga|Otsego|Tioga|Tompkins", replacement = "Central New York", borough)) %>% 
  transform(borough=gsub(pattern="Chemung|Genesee|Livingston|Monroe|Ontario|Schuyler|Seneca|Steuben|Wayne|Yates",replacement="Finger Lakes", borough)) %>% 
  transform(borough=gsub(pattern="Allegany|Cattaraugus|Chautauqua|Erie|Niagara|Orleans|Wyoming",replacement="Western New York", borough)) %>% 
  transform(borough=gsub(pattern="Delaware|Dutchess|Orange|Putnam|Sullivan|Ulster",replacement="Hudson Valley", borough)) %>% 
  transform(borough=gsub(pattern="Rockland|Westchester", replacement="Westchester/Rockland", borough)) %>% 
  transform(borough=gsub(pattern="Suffolk|Nassau", replacement="Long Island", borough)) %>% 
    transform(borough=gsub(pattern="Bronx|Brooklyn|Manhattan|Queens|Staten Island", replacement="New York City", borough)) %>% 
  group_by(borough) %>% 
  summarize (
    total_rate= sum(total), 
    Non_Hispanic_Count= sum(Non_Hispanic_Count),
    NH_White_Only_Count= sum(NH_White_Only_Count), 
    NH_Black_Only_Count= sum(NH_Black_Only_Count),
    NH_Other_Count = sum(NH_Other_Count))
```

renaming live birth H data of counties into regions.

```{r rename induced abortions}
rename_lb_h = clean_lb_h %>% 
  transform(borough = gsub(pattern = "Albany|Columbia|Fulton|Greene|Montgomery|Rensselaer|Saratoga|Schenectady|Schoharie|Warren|Washington", replacement = "Saratoga", borough)) %>%
  transform(borough = gsub(pattern = "Franklin|Clinton|Essex|Hamilton|Essex/Hamilton|Hamilton/Essex", replacement = "Eastern Adirondacks", borough)) %>%
  transform(borough = gsub(pattern = "Herkimer|Jefferson|Lewis|Oneida|Oswego|St Lawrence", replacement = "Western Adirondacks", borough)) %>%
  transform(borough = gsub(pattern = "Broome|Cayuga|Chenango|Cortland|Madison|Onondaga|Otsego|Tioga|Tompkins", replacement = "Central New York", borough)) %>% 
  transform(borough=gsub(pattern="Chemung|Genesee|Livingston|Monroe|Ontario|Schuyler|Seneca|Steuben|Wayne|Yates",replacement="Finger Lakes", borough)) %>% 
  transform(borough=gsub(pattern="Allegany|Cattaraugus|Chautauqua|Erie|Niagara|Orleans|Wyoming",replacement="Western New York", borough)) %>% 
  transform(borough=gsub(pattern="Delaware|Dutchess|Orange|Putnam|Sullivan|Ulster",replacement="Hudson Valley", borough)) %>% 
  transform(borough=gsub(pattern="Rockland|Westchester", replacement="Westchester/Rockland", borough)) %>% 
  transform(borough=gsub(pattern="Suffolk|Nassau", replacement="Long Island", borough)) %>% 
    transform(borough=gsub(pattern="Bronx|Brooklyn|Manhattan|Queens|Staten Island", replacement="New York City", borough)) %>% 
  group_by(borough) %>% 
  summarize (
    total_rate= sum(total), 
    Hispanic_Count= sum(Hispanic_Count),
    H_White_Only_Count= sum(H_White_Only_Count), 
    H_Black_Only_Count= sum(H_Black_Only_Count))
```

renaming live birth NH data of counties into regions.

```{r rename abortion NH}
rename_lb_nh= clean_lb_nh %>% 
  transform(borough = gsub(pattern = "Albany|Columbia|Fulton|Greene|Montgomery|Rensselaer|Saratoga|Schenectady|Schoharie|Warren|Washington", replacement = "Saratoga", borough)) %>%
  transform(borough = gsub(pattern = "Franklin|Clinton|Essex|Hamilton|Essex/Hamilton|Hamilton/Essex", replacement = "Eastern Adirondacks", borough)) %>%
  transform(borough = gsub(pattern = "Herkimer|Jefferson|Lewis|Oneida|Oswego|St Lawrence", replacement = "Western Adirondacks", borough)) %>%
  transform(borough = gsub(pattern = "Broome|Cayuga|Chenango|Cortland|Madison|Onondaga|Otsego|Tioga|Tompkins", replacement = "Central New York", borough)) %>% 
  transform(borough=gsub(pattern="Chemung|Genesee|Livingston|Monroe|Ontario|Schuyler|Seneca|Steuben|Wayne|Yates",replacement="Finger Lakes", borough)) %>% 
  transform(borough=gsub(pattern="Allegany|Cattaraugus|Chautauqua|Erie|Niagara|Orleans|Wyoming",replacement="Western New York", borough)) %>% 
  transform(borough=gsub(pattern="Delaware|Dutchess|Orange|Putnam|Sullivan|Ulster",replacement="Hudson Valley", borough)) %>% 
  transform(borough=gsub(pattern="Rockland|Westchester", replacement="Westchester/Rockland", borough)) %>% 
  transform(borough=gsub(pattern="Suffolk|Nassau", replacement="Long Island", borough)) %>% 
    transform(borough=gsub(pattern="Bronx|Brooklyn|Manhattan|Queens|Staten Island", replacement="New York City", borough)) %>% 
  group_by(borough) %>% 
  summarize (
    total_rate= sum(total), 
    Non_Hispanic_Count= sum(Non_Hispanic_Count),
    NH_White_Only_Count= sum(NH_White_Only_Count), 
    NH_Black_Only_Count= sum(NH_Black_Only_Count),
    NH_Other_Count = sum(NH_Other_Count))
```

#### Merging

merging data to get rates from count data

```{r merge_h}
merged_data_h =
  full_join(rename_lb_h, rename_ia_h, by = "borough") %>%
  janitor::clean_names() %>%
  mutate(
    hispanic_count = (hispanic_count_y / hispanic_count_x)*1000,
    h_white_only_count = (h_white_only_count_y / h_white_only_count_x)*1000,
    h_black_only_count = (h_black_only_count_y / h_black_only_count_x)*1000,
    total = (total_rate_y / total_rate_x)*1000) %>% 
select(borough, hispanic_count, h_white_only_count, h_black_only_count, total)

write_csv(merged_data_h, file = "data/NY_race_h.csv")

```

```{r merge_nh}
merged_data_nh =
  full_join(rename_lb_nh, rename_ia_nh, by = "borough") %>%
  janitor::clean_names() %>%
  mutate(
    non_hispanic_count = (non_hispanic_count_y / non_hispanic_count_x)*1000,
    nh_white_only_count = (nh_white_only_count_y / nh_white_only_count_x)*1000,
    nh_black_only_count = (nh_black_only_count_y / nh_black_only_count_x)*1000,
    nh_other_count = (nh_other_count_y / nh_other_count_x)*1000,
    total = (total_rate_y / total_rate_x)*1000) %>% 
select(borough, non_hispanic_count, nh_white_only_count, nh_black_only_count, nh_other_count, total)

write_csv(merged_data_nh, file = "data/NY_race_nh.csv")

```

#### Plotting

ggplot: induced abortion rates vs financial plans

```{r viz_finance}
abortion_race_hplot = 
  merged_data_h %>%
  select(-total) %>%
  pivot_longer(
    hispanic_count:h_black_only_count,
    names_to = "race", 
    values_to = "abortion"
  ) %>%
plot_ly(x = ~race, y = ~abortion, color = ~borough, type = "bar", colors = "viridis") %>%   layout(title = 'Abortion Ratios by Hispanic Race for NY Sate', yaxis = list(title = 'Number of Induced Abortions per 1,000 Live Births'))

abortion_race_hplot
```

```{r}
abortion_race_nhplot = 
  merged_data_nh %>%
  select(-total) %>%
  pivot_longer(
    non_hispanic_count:nh_other_count,
    names_to = "race", 
    values_to = "abortion"
  ) %>%
plot_ly(x = ~race, y = ~abortion, color = ~borough, type = "bar", colors = "viridis") %>%   layout(title = 'Abortion Ratios by Non-Hispanic Race for Boroughs and New York State', yaxis = list(title = 'Number of Induced Abortions per 1,000 Live Births'))

abortion_race_nhplot
```

